---
title: "Exploratory Data Analysis for SPROJ"
author: "Leonardo Andrade"
format: 
  pdf:
    warning: false
    message: false
editor: visual
---

# PACKAGES

In order to do this Exploratory Data Analysis, I will use the following packages:

```{r, load packages, warning= FALSE}
library(ggplot2) # for plotting data
library(readxl) # for reading excel files if necessary
library(dplyr) # for data wrangling
# library(arrow)
library(janitor) # for cleaning messy names of datasets
library(readr) # for reading files
library(gt) # for making 'publication ready' tables
library(usethis) #enables git pushing to put on github
library(gitcreds)
library(purrr) # this package is to loop and automate this entire process, instead of going through each sector 01 to 99
library(fixest) #fixed fx
library(broom)
library(haven) ## allows to read .dta files meant for stata
library(tidyverse)
library(tidysynth)
library(gsynth)
```

This senior project document will serve as the backbone of methodology and analysis part of the Senior Project"

# Motivation

According to the UN Trade Hub and Development:

(https://unctadstat.unctad.org/datacentre/reportInfo/US.RCA)

"Revealed comparative advantage (RCA) is based on Ricardian trade theory, which posits that patterns of trade among countries are governed by their relative differences in productivity. Although such productivity differences are difficult to observe, an RCA metric can be readily calculated using trade data to "reveal" such differences. While the metric can be used to provide a general indication and first approximation of a country's competitive export strengths, it should be noted that applied national measures which affect competitiveness such as tariffs, non-tariff measures, subsidies and others are not taken into account in the RCA metric."

"When a country has a revealed comparative advantage for a given product (RCA \>1), it is inferred to be a competitive producer and exporter of that product relative to a country producing and exporting that good at or below the world average. A country with a revealed comparative advantage in product i is considered to have an export strength in that product. The higher the value of a country’s RCA for product i, the higher its export strength in product i."

Part of this senior project aims to explain shifts in Revealed Comparative Advantage through a political economic lens, specifically in Mexico's Agriculture industry As the UN mentions, this RCA is merely an inference variable, and is used as a 'first approximation' in comparative advantage. Many of the shifts in RCA over time are not captured. I will use classical economic theory, neoclassical theory, economic literature, and papers to describe shifts in the RCA through a political economic lens. My sector analysis in these documents will be done in order of sector code and number, starting from 01; Animals Live, throughout the rest of the agricultural sector.

The agricultural sector is one that is extremely broad- ranging from plants and livestock, to coffee beans and even honey. In order to further analyze the RCA index, we must discern what codes constitute the agriculture industry. Each sector has its own general product code at the 2 digit, 4 digit, and 6 digit level, with each digits being more specific in each product. Using the International Trade Administration (ITA) Harmonized System (HS) codes, we will use these definitions to discern what products are part of the agriculture industry. There is opinions in the literature that showws variations in the product codes that is part of the 'agricultural sector.' I will use the definitions from this paper (https://www.researchgate.net/publication/332551060_Changes_in_Foreign_Trade_in_Agricultural_Products_between_China_and_Poland) to classify the codes that belongs in the agricultural sector. Furthermore, I will analyze each sector at a general level, starting at the 2 digit, and then up to 4 digits. Due to constraints and limitations of the publicly available data, the maximum scope of analysis is up to the 4 digit HS code level. The data will also be split up into multiple time periods. Period A will be from 1992-2000, period B will be from 2000-2008, and period C will be from 2008-2016, period D will be from 2016-2024.


# Functions

The section below describes the functions which are necessary in computing this analysis, I have made comments in the code chunks in order to further expand on each line of code and what it does.

### Necessary Variables Function

```{r, arrange times}
## This is a function that takes in only the necessary variables for an RCA analysis, including arranging times in order from least to greatest

necessary_variables <- function(df) { # inputs a dataframe
  df |> 
    dplyr::mutate(cmdCode = as.character(cmdCode)) |>
    arrange(period) |> #arranges the df by year
    select(period, reporterISO, cmdCode, flowCode, qty, netWgt, fobvalue, primaryValue) #selects the variables necessary for RCA
}
```

### RCA Function

```{r, RCA calculation}
## RCA calculation function

#inputs country exports of product i , country's total exports, world exports of product i, and world total exports
calculate_rca <- function(country_product, country_total, world_product, world_total) {
  
  #join by period to align years
  rca_df <- country_product |>
    inner_join(country_total, by = "period", suffix = c("_prod", "_total")) |>
    inner_join(world_product, by = "period") |>
    inner_join(world_total, by = "period", suffix = c("_world_prod", "_world_total")) |>
    mutate(
      RCA = (fobvalue_prod / fobvalue_total) / 
            (fobvalue_world_prod / fobvalue_world_total)
    ) |>
    select(period, RCA)
  
  return(rca_df)
}

```

### generalized RCA Function

```{r}
calculate_rca_generalized <- function(country_sector_df,
                                      world_sector_df,
                                      country_all_df,
                                      world_all_df) {
  
  #first, I aggregate by year and product, with variables period, and cmdCode, removing any na's
  country_prod <- country_sector_df |>
    group_by(period, cmdCode) |>
    summarise(country_product_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")
  
  world_prod <- world_sector_df |>
    group_by(period, cmdCode) |>
    summarise(world_product_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")
  
  #next, I aggregate total exports using fobvalue  (denominator)
  country_all <- country_all_df |>
    group_by(period) |>
    summarise(total_country_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")
  
  world_all <- world_all_df |>
    group_by(period) |>
    summarise(total_world_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")
  
  # 3. next, I merge everytjing by year and commodity code
  rca_df <- country_prod |>
    left_join(world_prod, by = c("period", "cmdCode")) |>
    left_join(country_all, by = "period") |>
    left_join(world_all, by = "period") |>
    mutate(
      RCA = (country_product_value / total_country_value) /
            (world_product_value / total_world_value),
      has_CA = RCA >= 1
    ) |>
    arrange(period, cmdCode)
  
  return(rca_df) #outputs a dataframe 
}
```




# Recocded loading dataframes

```{r}

#setup 

base_path <- "~/Desktop/SPROJ/workplace/input"

time_periods <- c("a", "b", "c")

sector_codes <- c("01", "02", "04", "05", "07", "08", "09", "10", "11",
                  "12", "13", "14", "15", "16", "17", "18", "19", "20",
                  "21", "22", "23", "24", "50", "51", "52", "53")


# functions (unchanged from your original)

necessary_variables <- function(df) {
  df |>
    dplyr::mutate(cmdCode = as.character(cmdCode)) |>
    arrange(period) |>
    select(period, reporterISO, cmdCode, flowCode, qty, netWgt, fobvalue, primaryValue)
}

calculate_rca_generalized <- function(country_sector_df,
                                      world_sector_df,
                                      country_all_df,
                                      world_all_df) {
  country_prod <- country_sector_df |>
    group_by(period, cmdCode) |>
    summarise(country_product_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")

  world_prod <- world_sector_df |>
    group_by(period, cmdCode) |>
    summarise(world_product_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")

  country_all <- country_all_df |>
    group_by(period) |>
    summarise(total_country_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")

  world_all <- world_all_df |>
    group_by(period) |>
    summarise(total_world_value = sum(fobvalue, na.rm = TRUE), .groups = "drop")

  rca_df <- country_prod |>
    left_join(world_prod, by = c("period", "cmdCode")) |>
    left_join(country_all, by = "period") |>
    left_join(world_all, by = "period") |>
    mutate(
      RCA = (country_product_value / total_country_value) /
            (world_product_value / total_world_value),
      has_CA = RCA >= 1
    ) |>
    arrange(period, cmdCode)

  return(rca_df)
}


#LOAD AND COMBINE TOTAL EXPORT FILES ACROSS TIME PERIODS


load_combined_totals <- function(folder_prefix, file_prefix, time_periods, base_path) {
  # Loops over time periods a, b, c and stacks into one dataframe
  map_dfr(time_periods, function(t) {
    file_path <- file.path(base_path,
                           paste0(folder_prefix, t), # i.e mx_total_xports_world_time_a
                           paste0(file_prefix, t, ".csv")) # e.x mx_xports_all_time_a.csv
    if (!file.exists(file_path)) {
      message("File not found, skipping: ", file_path)
      return(NULL)
    }
    read_csv(file_path, show_col_types = FALSE) |> necessary_variables()
  })
}

# Combined Mexico total exports: 1990-2012
mx_xports_all <- load_combined_totals(
  folder_prefix = "mx_total_xports_world_time_",
  file_prefix   = "mx_xports_all_time_",
  time_periods  = time_periods,
  base_path     = base_path
)

# Combined world total exports: 1990-2012
world_xports_all <- load_combined_totals(
  folder_prefix = "world_total_xports_time_",
  file_prefix   = "world_xports_all_time_",
  time_periods  = time_periods,
  base_path     = base_path
)


# PROCESS SECTOR ACROSS ALL THREE TIME PERIODS

process_sector <- function(sector) {
  sector_data <- map_dfr(time_periods, function(t) {

    # mx files: try short name first (time_a and time_c pattern)
    mx_file <- file.path(base_path,
                         paste0("mx_xports_prod_i_time_", t),
                         paste0("mx_sector_", sector, "_time_", t, ".csv"))

    # if not found, try long name (time_b pattern and sector 01)
    if (!file.exists(mx_file)) {
      mx_file <- file.path(base_path,
                           paste0("mx_xports_prod_i_time_", t),
                           paste0("mx_xports_sector_", sector, "_time_", t, ".csv"))
    }

    # world files: always world_xports_sector_XX_time_Y.csv
    world_file <- file.path(base_path,
                            paste0("world_xports_prod_i_time_", t),
                            paste0("world_xports_sector_", sector, "_time_", t, ".csv"))

    if (!file.exists(mx_file) | !file.exists(world_file)) {
      message("Skipping sector ", sector, " time_", t, " (file not found)")
      return(NULL)
    }

    mx_sector <- read_csv(mx_file, show_col_types = FALSE) |> 
      necessary_variables()
    
    world_sector <- read_csv(world_file, show_col_types = FALSE) |> necessary_variables()

    rca_slice <- calculate_rca_generalized(
      mx_sector,
      world_sector,
      mx_xports_all,
      world_xports_all
    )

    return(rca_slice)
  })

  if (is.null(sector_data) | nrow(sector_data) == 0) return(NULL)
  sector_data$sector <- sector
  return(sector_data)
}


# BUILD THE BIG RCA DATAFRAME


dir.create("output/clean/rca_sectors", recursive = TRUE, showWarnings = FALSE)

rca_big_df <- map_dfr(sector_codes, function(s) {
  result <- process_sector(s)
  if (!is.null(result)) {
    write_csv(result, paste0("output/clean/rca_sectors/rca_sector_", s, ".csv"))
    message("Saved sector ", s)
  }
  return(result)
})

# Add sector column and NAFTA dummies
rca_big_df <- rca_big_df |>
  mutate(
    cmdCode = as.character(cmdCode),
    sector = substr(cmdCode, 1, 2),
    post_nafta = if_else(period >= 1994, 1, 0),
    nafta_phase = case_when(
      period < 1994              ~ "pre",
      period >= 1994 & period <= 2000 ~ "post"
    )
  )
```


# SECTORS

## 01 Animals; Live

This section will calculate RCA over time before NAFTA of the animal section, containing HS product code of 0101 to 0106.

```{r}
## load the datasets
rca_sector_01 <- rca_big_df |>
  filter(sector == "01", nchar(cmdCode) == 4)
```



```{r}
ggplot(rca_sector_01, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") + # NAFTA line
  annotate("text", x = 1994.3, y = max(rca_sector_01$RCA, na.rm = TRUE),
           label = "NAFTA", hjust = 0, size = 3, color = "blue") +
  labs(title = "RCA Over Time – Sector 01: Animals Live",
       subtitle = "1990–2012",
       x = "Year", y = "RCA",
       color = "HS4 Code") +
  theme_bw()
```

Interesting that Mexico seems to have a general RCA of Live Animals, but there was a sharop decline starting in 1993. NAFTA was enacted January 1, 1994- was it a driving factor in RCA shift? What about data from future years?

## 02 Meat and Edible Meat offal

```{r}
## load the datasets
rca_sector_02 <- rca_big_df |>
  filter(sector == "02", nchar(cmdCode) == 4)

```

```{r}
## graph RCA
ggplot(rca_sector_02, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") + # NAFTA line
  annotate("text", x = 1994.3, y = max(rca_sector_01$RCA, na.rm = TRUE),
           label = "NAFTA", hjust = 0, size = 3, color = "blue") +
  labs(title = "RCA Over Time – Sector 02: Meat and Edible Meat offal",
       subtitle = "1990–2012",
       x = "Year", y = "RCA",
       color = "HS4 Code") +
  theme_bw()
```

Sector of 02 seems to not have much of a comparative advantage in Mexico.

## 04 Dairy Produce; Birds' eggs; Natural Honey; Edible products of animal origin

- Omit 0409: Honey.

```{r}
rca_sector_04 <- rca_big_df |>
  filter(sector == "04", nchar(cmdCode) == 4)
```

Graph Sector 04

```{r}
## graph RCA
ggplot(rca_sector_04, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 2)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 04: Dairy Produce;",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 05 Oher Animal Products

ANIMAL ORIGINATED PRODUCTS; NOT ELSEWHERE SPECIFIED OR INCLUDED

We omit the following

-   0501 Human hair; unworked, whether or not washed or scoured; waste of human hair

-   0509 Sponges

-   0511 Animal products not elsewhere specified or included; dead animals of chapter 1 or 3, unfit for human consumption

```{r}
rca_sector_05 <- rca_big_df |>
  filter(sector == "05", nchar(cmdCode) == 4)
```

```{r}
ggplot(rca_sector_05, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 30)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 05: Other Animal Products",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

Interestingly, 0503 "Horsehair, waste" has a comparative advantage in that product, but the sector does not have a comparative advantage.

## 07- Vegetables and Certain Roots and Tubes; Edible

```{r}
rca_sector_07 <- rca_big_df |>
  filter(sector == "07", nchar(cmdCode) == 4)
```

```{r}
ggplot(rca_sector_07, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 30)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 07: Vegetables and Certain Roots and Tubes; Edible",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

\textbf{Sector Analysis : 07}

Now, I are tasked to clean the dataset and calculate an RCA given the dataset with the entire sector. Using the code above, find a way to generalize the RCA calculation to continue the calculation, with an ultimate goal of showing an RCA graph for each product in the sector.


Now, I must use the necessary variables for the RCA. necessary variables are the following: period, reporterISO, cmdCode, flowCode, qty, netWgt, fobvalue, primaryValue. Recall that the RCA variable is calculated with the following:

RCA = \[(Country Exports of Product i / Country A Exports of ALL products) / (World Exports of product i / World Exports of all products)\]

$\text{RCA}_i = \frac{\dfrac{X_{A,i}}{X_{A,\text{all}}}} {\dfrac{X_{World,i}}{X_{World,\text{all}}}}$

The RCA variable calculates a share of country exports a product over the world exports of said product. If the RCA variable is greater than, or equal to 1, then the country has a revealed comparative advantage in producing said good. According to neoclassical theory, gains from trade are still probable in comparative advantage. Theory says to export goods in which the country has a comparative advantage in producing.

Now, I have the necessary variables, Country MX Exports of product 0701 -\> 0710. I have the data of Mexico's exports of all products, titled `mx_xports_all`. Now, I need World's exports of products 0701 -\> 0710. I can do this by using UNCOMTRADE data as Ill. Repeating the same steps, I will grab the necessary variables.


I now have World exports of product 0701 -\> 0710 with the same time period of 1992 to 2000. Recall I have datasets of Mexico's exports of all products titled `mx_xports_all` with variables period, reporterISO, cmdCode, flowCode, qty, netWgt, fobvalue, primaryValue. I also have `world_xports_all` which contains world exports of all products with variables period, reporterISO, cmdCode, flowCode, qty, netWgt, fobvalue, primaryValue. HoIver, notice that `world_xports_sector_07` and `mx_xports_sector_07` have the cmdCode, hoIver, they are not aggregated, and their cmdCode varies from 0701 -\> 0710. Below is an attempted geenralized calclation of RCA, with a clear difference of grouping by `group_by(period, cmdCode)` due to the nature of the dataframe with varying cmdCode

Let's test this in action to see if it works!

RCA of sector 0701 to 0710.


Visualize over time:

```{r}
## ggplot to bisualize over time
ggplot(rca_sector_07, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "RCA Over Time – HS07 Sector Products (0701–0710)",
       x = "Year", y = "RCA") +
  theme_bw()
```

I can now develop code to show which products have an RCA \> 1 and RCA \< 1.

I can do this by running a simple filter:

```{r}
## show which years had an RCA > 1
table_RCA_gt1 <- rca_sector_07 |>
  filter(RCA > 1) |>
  arrange(period, cmdCode)
table_RCA_gt1
```

or by showing which years had an RCA \> 1

```{r}
#summarize the comparative advantage by product by grouping by product code
summary_CA_by_product <- rca_sector_07 |>
  group_by(cmdCode) |>
  summarise(
    years_with_CA = paste(period[RCA > 1], collapse = ", "),
    n_years_CA = sum(RCA > 1),
    total_years = n(),
    share_years_CA = n_years_CA / total_years
  )
```

and a publication ready table:

```{r}
summary_CA_by_product |> 
  gt() |>
  tab_header(
    title = "Mexico’s RCA, Sector 07 (RCA > 1)",
  )
```

I can fix this dataframe by changing the name of the previous one

\textbf{Sector definitions}

07 - VEGETABLES AND CERTAIN ROOTS AND TUBERS; EDIBLE

-   0701 - Potatoes; fresh or chilled

-   0702 - Tomatoes; fresh or chilled

-   0703 - Onions, shallots, garlic, leeks and other alliaceous vegetables; fresh or chilled

-   0704 - Cabbages, caulifloIrs, kohlrabi, kale and similar edible brassicas; fresh or chilled

-   0705 - Lettuce (lactuca sativa) and chicory (cichorium spp.) fresh or chilled

-   0706 - Carrots, turnips, salad beetroot, salsify, celeriac, radishes and similar edible roots; fresh or chilled

-   0707 - Cucumbers and gherkins; fresh or chilled

-   0708 - Leguminous vegetables; shelled or unshelled, fresh or chilled

-   0709 - Vegetables; n.e.c. in chapter 07, fresh or chilled

-   0710 - Vegetables (uncooked or cooked by steaming or boiling in water); frozen

-   0711 - Vegetables provisionally preserved; (e.g. by sulphur dioxide gas, in brine, in sulphur water or in other preservative solutions), but unsuitable in that state for immediate consumption

-   0712 - Vegetables, dried; whole, cut, sliced, broken or in powder, but not further prepared

-   0713 - Vegetables, leguminous; shelled, whether or not skinned or split, dried

## 08- FRUIT AND NUTS, EDIBLE; PEEL OF CITRUS FRUIT OR MELONS

We omit the following:

-   0812 Fruit and nuts provisionally preserved; e.g. by sulphur dioxide gas, brine, in sulphur water or in other preservative solutions, but unsuitable in that state for immediate consumption

-   0813 Fruit, dried, other than that of heading no. 0801 to 0806; mixtures of nuts or dried fruits of this chapter

-   0814 Peel of citrus fruit or melons (including watermelons); fresh, frozen dried or provisionally preserved in brine, in sulphur water or in other preservative solutions


```{r}
# label: debug-check
exists("rca_big_df")
nrow(rca_big_df)
```


```{r}
rca_sector_08 <- rca_big_df |>
  filter(sector == "08", nchar(cmdCode) == 4)
```

```{r}
ggplot(rca_sector_08, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 30)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 08: FRUIT AND NUTS, EDIBLE; PEEL OF CITRUS FRUIT OR MELONS",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```


## 09- COFFEE, TEA, MATE AND SPICES

No 4 digit HS classification code is omitted.

```{r}
rca_sector_09 <- rca_big_df |>
  filter(sector == "09", nchar(cmdCode) == 4)
```


Plot sector 09

```{r}
ggplot(rca_sector_09, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 09: COFFEE, TEA, MATE AND SPICES",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 10 CEREALS 

No 4 digit HS classification code is omitted.

```{r}
rca_sector_10 <- rca_big_df |>
  filter(sector == "10", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_10, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 10: CEREALS",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 11 Milling products; malt; starch, etc.; gluten

No 4 digit HS classification code is omitted.

No 4 digit HS classification code is omitted.

```{r}
rca_sector_11 <- rca_big_df |>
  filter(sector == "11", nchar(cmdCode) == 4)
```


Plot sector 11

```{r}
ggplot(rca_sector_11, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 11: Milling products; malt; starch, etc.; gluten",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 12 Oilseeds; kernels; industrial medicinal plants; feed

We omit the following HS Codes:

-   1205 Rape or colza seeds; whether or not broken

-   1211 Plants and parts of plants (including seeds and fruits), of a kind used primarily in perfumery, in pharmacy or for insecticidal, fungicidal or similar purposes, fresh, chilled, frozen or dried, whether or not cut, crushed or powdered

-   1212 Locust beans, seaweeds and other algae, sugar beet, sugar cane, fresh, chilled, frozen or dried, whether or not ground; fruit stones, kernels and other vegetable products (including unroasted chicory roots) used primarily for human consumption, n.e.c.

-   1214 Swedes, mangolds, fodder roots, hay, lucerne (alfalfa), clover, sainfoin, forage kale, lupines, vetches and similar forage products, whether or not in the form of pellets

```{r}
rca_sector_12 <- rca_big_df |>
  filter(sector == "12", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_12, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 12: Oilseeds; kernels; industrial medicinal plants; feed",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 13 Shellac; gum, fat and other plant liquids, juice

```{r}
rca_sector_13 <- rca_big_df |>
  filter(sector == "13", nchar(cmdCode) == 4)
```


Plot sector 13

```{r}
ggplot(rca_sector_13, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 13: Shellac; gum, fat and other plant liquids, juice",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 14 Planting materials for planting; other plant products

```{r}
rca_sector_14 <- rca_big_df |>
  filter(sector == "14", nchar(cmdCode) == 4)
```


Plot sector 14

```{r}
ggplot(rca_sector_14, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 14: Planting materials for planting; other plant products",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 15 Animal and vegetable oils and fats; refined edible oils and fats

We Omit the following:

-   1503

-   1505

-   1506

-   1508

-   1510

-   1511 to 1514

-   1517 to 1519

```{r}
rca_sector_15 <- rca_big_df |>
  filter(sector == "15", nchar(cmdCode) == 4)
```


Plot sector 15

```{r}
ggplot(rca_sector_15, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 15: Animal and vegetable oils and fats...",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 16 Meat, fish and other aquatic invertebrate products

RCA Sector 16

```{r}
rca_sector_16 <- rca_big_df |>
  filter(sector == "16", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_16, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 16: Meat, fish and other aquatic invertebrate products",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 17 Sugar and confectionery

RCA Sector 17

```{r}
rca_sector_17 <- rca_big_df |>
  filter(sector == "17", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_17, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 17:  Sugar and confectionery",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 18 Cocoa and cocoa products

RCA Sector 18

```{r}
rca_sector_18 <- rca_big_df |>
  filter(sector == "18", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_18, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 18: Cocoa and cocoa products",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 19 Cereal flour, starch, etc. or dairy products; cakes

RCA Sector 19

No HS Code is omitted
```{r}
rca_sector_19 <- rca_big_df |>
  filter(sector == "19", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_19, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 19: Cereal flour, starch, etc. or dairy products; cakes",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 20 Products of vegetables, fruits or other parts of plants

RCA Sector 20

No HS Code is omitted

```{r}
rca_sector_20 <- rca_big_df |>
  filter(sector == "20", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_20, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 43)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 20: Products of vegetables, fruits or other parts of plants",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 21 Miscellaneous food

RCA Sector 21

No HS Code is omitted

```{r}
rca_sector_21 <- rca_big_df |>
  filter(sector == "21", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_21, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Miscellaneous food",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```


## 22 Beverages, wine and vinegar

RCA Sector 22

No HS Code is omitted

```{r}
rca_sector_22 <- rca_big_df |>
  filter(sector == "22", nchar(cmdCode) == 4)
```


Plot sector 22

```{r}
ggplot(rca_sector_22, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 22 Beverages, wine and vinegar",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 23 Food industry residues and waste; formulated feed

RCA Sector 23

No HS Code is omitted

```{r}
rca_sector_23 <- rca_big_df |>
  filter(sector == "23", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_23, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 2)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 23: Food industry residues and waste; formulated feed",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 24 Tobacco, tobacco and tobacco substitute products

RCA Sector 24

No HS Code is omitted.


```{r}
rca_sector_24 <- rca_big_df |>
  filter(sector == "24", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_24, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 1.5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 24: Tobacco, tobacco and tobacco substitute products",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 50 Silk

RCA Sector 50

No HS Code is omitted

```{r}
rca_sector_50 <- rca_big_df |>
  filter(sector == "50", nchar(cmdCode) == 4)
```


Plot sector 50

```{r}
ggplot(rca_sector_50, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 50: Silk",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 51 Wool and other animal hair

RCA Sector 51

We omit the following HS Codes

-   5103 Waste of wool or of fine or coarse animal hair, including yarn waste but excluding garnetted stock\

-   5104 Wool, or fine or coarse animal hair; garnetted stock

-   5113 Woven fabrics of coarse animal hair or of horsehair

```{r}
rca_sector_51 <- rca_big_df |>
  filter(sector == "51", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_51, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 2)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 51: Wool and other animal hair",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

## 52 Cotton

Omit

-   5205 Cotton yarn (other than sewing thread), containing 85% or more by weight of cotton, not put up for retail sale

-   5206 Cotton yarn (other than sewing thread), containing less than 85% by weight of cotton, not put up for retail sale

```{r}
rca_sector_52 <- rca_big_df |>
  filter(sector == "52", nchar(cmdCode) == 4)
```


Plot sector 10

```{r}
ggplot(rca_sector_52, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 5)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 52: Cotton",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```


## 53 Other plant fibers

We omit the following:

-   5310 Woven fabrics of jute, other textile bast fibres of heading no. 5303

-   5311 Woven fabrics of other vegetable textile fibres; woven fabrics of paper yarn


```{r}
rca_sector_53 <- rca_big_df |>
  filter(sector == "53", nchar(cmdCode) == 4)
```


Plot sector 53

```{r}
ggplot(rca_sector_53, aes(x = period, y = RCA, color = cmdCode, group = cmdCode)) +
  geom_line(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1994, linetype = "dotted", color = "blue") +
  coord_cartesian(ylim = c(0, 4.2)) + # cap at 10, adjust as needed
  labs(title = "RCA Over Time – Sector 53: Other plant fibers",
       subtitle = "1990–2012", x = "Year", y = "RCA", color = "HS4 Code") +
  theme_bw()
```

# Load Variables for Regression


```{r}
# load exchange rate dataset and tidy up
exch_rate <- read_csv("regression/variables/exch_rates.csv")

# (https://databank.worldbank.org/reports.aspx?source=2&series=PA.NUS.FCRF&country=MEX)


# load dataset for fdi inflow and tidy up

fdi_inflow <- read_csv("regression/variables/fdi_inflow_to_mx.csv") # worldbank filtered out 

#fdi_inflow <- fdi_inflow |>
  #  select(Time, `Foreign direct investment, net inflows (BoP, current US$) [BX.KLT.DINV.CD.WD]`) |>
  #rename(period = "Time", fdi_inflow_usd = `Foreign direct investment, net inflows (BoP, current US$) [BX.KLT.DINV.CD.WD]`) |>
 # na.omit(fdi_inflow)


# load datasets for agriculture fdi

agriculture_fdi_pt_1 <- read_csv("regression/variables/agriculture_fdi_pt_1.csv") ## https://www.gob.mx/se/acciones-y-programas/competitividad-y-normatividad-inversion-extranjera-directa

fdi_agri_mexico_2 <- read_csv("regression/variables/fdi_agri_mexico_2.csv")

agriculture_fdi_pt_1 <- agriculture_fdi_pt_1 |>
  select(period, agri_fdi)


agriculture_fdi_pt_2 <- fdi_agri_mexico_2 |>
  filter(anio >= 2001 & anio <= 2012) |>
  mutate(fn_millones_de_dolares = as.numeric(fn_millones_de_dolares)) |>
  group_by(anio) |>
  summarise(total_agri_fdi = sum(fn_millones_de_dolares, na.rm = TRUE), .groups = "drop") |>
  rename(period = anio)

agriculture_fdi_total <- agriculture_fdi_pt_1 |>
  mutate(agri_fdi = agri_fdi / 1e6) |>
  rename(total_agri_fdi = agri_fdi) |>
  bind_rows(agriculture_fdi_pt_2) |>
  arrange(period)

agriculture_fdi_total

```

# Reg/EDA Df

```{r}

## Create main Dataframe
rca_panel <- rca_big_df |>
  left_join(fdi_inflow, by = "period") |>
  left_join(exch_rate, by = "period") |>
  left_join(agriculture_fdi_total, by = "period") |>
  filter(nchar(cmdCode) == 4) |>
  mutate(log_rca = log(RCA))
  

## store all cmdCodes as a vector

hs4_codes <- rca_panel |>
  pull(cmdCode) 
```


# EDA

Now that we have our main dataframe `rca_panel`, we can conduct an EDA to see whether or not we need to do any transformations before we model.

## 1) Distribution

```{r}
rca_panel |>
  select(RCA) |>
  ggplot(aes(x = RCA)) +
  geom_histogram(binwidth = 0.3) +
  theme_minimal()
```

We observe that our RCA is heavily right skewed. Hence, we *might* have to do a log(RCA) transformation due to the nature of the distribution.

## 2) Time

Show the time frame (1992-2000)

```{r}
rca_panel|>
  distinct(period) |> 
  arrange(period)
```

## 3) Num of products

```{r}
rca_panel |>
  distinct(cmdCode) |> #gives us distinct values for cmdCode
  nrow() #gives us the number of rows 
```

The number of HSCodes that we are analyzing is 194.

## 4) Balanced

Does every product appear in every year?

```{r}
rca_panel |>
  count(cmdCode) |>
  summary()
```

## 5) Correlations

We must check for correlations between our explanatory variables

```{r}
rca_panel |>
  distinct(period, post_nafta, exch_rate, fdi_inflow_usd, total_agri_fdi) |>
  select(post_nafta, exch_rate, total_agri_fdi )|>
  cor()
```

## 6) Transform the variables

```{r}
rca_panel <- rca_panel|>
  mutate(
    log_rca = log(RCA),  # if RCA is right-skewed
    log_exch = log(exch_rate),
    asinh_fdi = asinh(total_agri_fdi) # sinh = (e^x - e^-x)/2
  )
```


## 7) Num of Prods RCA > 1

```{r}
rca_greq1 <- rca_panel |>
  filter(RCA >= 1)

## what are the products with RCA >= 1? regardless of time

rca_greq1 |>
  select(cmdCode) |>
  unique()
```


## 8) Products which did not start off with comparative advantage before nafta, and ended up with one after?

```{r}

#products that did not have a CA before NAFTA
no_ca_before <- rca_panel |>
  filter(post_nafta == 0) |> ## before nafta
  group_by(cmdCode) |> #group by commodity Code
  summarise(had_ca = any(has_CA)) |> # if any values in has_CA is true, i.e does have a CA
  filter(had_ca == FALSE) |>
  pull(cmdCode)

no_ca_before

# of those products which ones developed a comparative advantage after 1994?
gained_ca_after <- rca_panel  |>
  filter(post_nafta == 1, cmdCode %in% no_ca_before) |>
  group_by(cmdCode) |>
  summarise(gained_ca = any(has_CA)) |>
  filter(gained_ca == TRUE) |>
  pull(cmdCode)

gained_ca_after
```

## 9) Descriptive overview

```{r}
rca_panel |>
  group_by(post_nafta) |>
  summarise(
    mean_rca = mean(RCA, na.rm = TRUE),
    median_rca = median(RCA, na.rm = TRUE),
    sd_rca = sd(RCA, na.rm = TRUE),
    min_rca = min(RCA, na.rm = TRUE),
    max_rca = max(RCA, na.rm = TRUE),
    pct_with_RCA = mean(has_CA)*100
  )
```

## 10) Prevalence in RCA by sector

```{r}
rca_panel |>
  group_by(sector, post_nafta) |>
  summarise(
    n_products= n_distinct(cmdCode), #counts num of distinct values
    pct_with_CA = mean(has_CA) * 100,
    mean_rca = mean(RCA, na.rm = TRUE)
  )
```

## 11) Year by Year

```{r}
rca_panel |>
  group_by(period) |>
  summarise(
    n_with_CA = sum(has_CA),
    total_products = n_distinct(cmdCode),
    pct_with_CA = mean(has_CA)*100
  )
```

## 12) Volatility of RCA

```{r}
 rca_panel |>
  arrange(cmdCode, period) |>
  group_by(cmdCode) |>
  summarise(
    switches = sum(has_CA != lag(has_CA), na.rm = TRUE)
  ) |>
  filter(switches > 1) |>
  arrange(desc(switches))
```

## 13 Regressor variation

```{r}
rca_panel |>
  group_by(nafta_phase) |>
  summarise(
    mean_exch_rate = mean(exch_rate, na.rm = TRUE),
    mean_fdi = mean(fdi_inflow_usd, na.rm = TRUE),
    mean_agri_fdi = mean(total_agri_fdi, na.rm = TRUE),
    n_obs = n()
  )
```

## 14) RCA Trajectories 

```{r}

get_rca_trajectory <- function(tempcmdcde) {
  rca_panel |>
    filter(cmdCode == tempcmdcde) |>
    select(period, cmdCode, RCA, has_CA, exch_rate, fdi_inflow_usd) |>
    arrange(period) |>
    mutate(rca_change = RCA - lag(RCA))  # how much RCA shifted each year
}

map(unique(rca_panel$cmdCode), get_rca_trajectory) |>
  list_rbind()
```

## 15) How long did it take for products to gain RCA after NAFTA?

```{r}
rca_panel |>
  filter(post_nafta == 1, has_CA == TRUE) |>
  group_by(cmdCode) |>
  summarise(first_ca_year = min(period)) |>
  mutate(years_after_nafta = first_ca_year - 1994) |>
  arrange(years_after_nafta) |>
  arrange(desc(years_after_nafta))
```

## 16) Consistency in RCAS

```{r}
rca_panel |>
  filter(post_nafta == 1) |>
  group_by(cmdCode) |>
  summarise(
    years_with_CA = sum(has_CA),
    total_years = n(),
    consistency = years_with_CA/total_years
  ) |>
  mutate(ca_type = case_when(
    consistency == 1 ~ "Persistent CA",
    consistency >= 0.5 ~ "Majority CA",
    consistency > 0 & consistency < 0.5 ~ "Sporadic CA",
    consistency == 0 ~ "No CA"
  )) |>
  count(ca_type)
```


## 17) Avg RCA By sector

```{r}
rca_panel |>
  group_by(sector, post_nafta) |>
  summarise(
    mean_rca = mean(RCA),
    median_rca  = median(RCA),
    n_products = n_distinct(cmdCode)
  ) |>
  pivot_wider(
    names_from = post_nafta,
    values_from = c(mean_rca, median_rca, n_products),
    names_prefix = "post_"
  )
```

## 18) Largest RCA Gains?

```{r}
rca_panel |>
  group_by(sector, post_nafta) |>
  summarise(mean_rca = mean(RCA)) |>
  pivot_wider(names_from = post_nafta, values_from = mean_rca, names_prefix = "period_") |>
  mutate(rca_gain = period_1 - period_0) |>
  arrange(desc(rca_gain))
```

## 19) FDI over time

```{r}
ggplot(data = rca_panel, aes(x = period, y = fdi_inflow_usd)) +
  geom_line() +
  theme_minimal() +
  xlab("Time") +
  ylab("Foreign Direct Investment to Mexico")
```

```{r}
ggplot(data = rca_panel, aes(x = period, y = total_agri_fdi)) +
  geom_line() +
  theme_minimal() +
  xlab("Time") +
  ylab("Agriculture Foreign Direct Investment to Mexico")
```


# MODELS

## Model 1 Fixed Effects Regresission

```{r}
model <- feols(RCA ~ post_nafta + log(exch_rate) + log(fdi_inflow_usd) | cmdCode,
               data = rca_panel,
               cluster = ~cmdCode)

summary(model)
```



## Model 4 USE THIS

```{r}

model4 <- feols(log(RCA) ~ post_nafta + log(exch_rate) + log(total_agri_fdi)|cmdCode,
               data = rca_panel,
               cluster = ~cmdCode)

summary(model4)

```

### Model 4 Specification


$$
\log(\text{RCA}_{it}) = \beta_1 \text{PostNAFTA}_{t} + \beta_2 \log(\text{ExchRate}_{t}) + \beta_3 \log(\text{AgriFFDI}_{t}) + \alpha_i + \varepsilon_{it}
$$

where $i$ indexes HS4 product codes, $t$ indexes years, and:

- $\log(\text{RCA}_{it})$ is the log of the Balassa (1965) Revealed Comparative Advantage index for product $i$ in year $t$
- $\text{PostNAFTA}_{t}$ is a binary indicator equal to 1 for $t \geq 1994$ and 0 otherwise
- $\log(\text{ExchRate}_{t})$ is the log of the Mexican peso--US dollar exchange rate (pesos per USD)
- $\log(\text{AgriFFDI}_{t})$ is the log of total foreign direct investment inflows into Mexico's agricultural sector
- $\alpha_i$ are product-level fixed effects absorbing time-invariant heterogeneity across HS4 codes
- $\varepsilon_{it}$ is the idiosyncratic error term, with standard errors clustered at the $\alpha_i$ (cmdCode) level

The model is estimated via Ordinary Least Squares with product-level fixed effects using the `feols` function from the `fixest` package in R [@bergé2018]. 
The within-transformation removes $\alpha_i$, so $\beta_1$ identifies the average change in log(RCA) 
associated with NAFTA implementation, holding exchange rate and agricultural FDI constant, 
within product codes over time.



### Interpretation

The interpretation of model 4 will be as follows.

**Log(Exchange Rate)**

A 1 percent increase in the exchange rate is assoiated with a 0.43 percent increase in RCA (at the 1 percent level)

**nafta_phaseearly**

Log(RCA) was 0.41 (units) lower than in the pre-NAFTA baseline, holding exchange rate and agriculture_FDI constant (significant at the 5 percent level )

\*\* asinh(agri_fdi)\*\*

A rise in agriculture FDI is associated with a decrease in RCA (counterintuitive).

A 1 percent increase in FDI in Mexico's Agriculture Sector is linked with a -0.142 percent decrease in RCA, holding all else constant.

-   FDI may have been used to increase domestic production rather than building export competitiveness in the global market.

-   FDI may flow into sectors that lose comparative advantage.

-   Aggregate agriculture FDI may have been flowing disproportionately into subsectors (i.e fertilizer, etc..) so FDI is not actually reaching the products driving the RCA variation.


# TABLES NEEDS FIXING

```{r}

# Main Model Table (Claude AI)

model4_tidy <- broom::tidy(model4, conf.int = TRUE) |>
  mutate(
    term = case_when(
      term == "nafta_phaseearly" ~ "NAFTA Phase: Early",
      term == "nafta_phasepost"  ~ "NAFTA Phase: Post",
      term == "log(exch_rate)"   ~ "Log Exchange Rate",
      term == "asinh(agri_fdi)"  ~ "IHS Agricultural FDI",
      TRUE ~ term
    ),
    stars = case_when(
      p.value < 0.01 ~ "***",
      p.value < 0.05 ~ "**",
      p.value < 0.1  ~ "*",
      TRUE           ~ ""
    ),
    estimate_stars = paste0(round(estimate, 3), stars),
    std.error = paste0("(", round(std.error, 3), ")")
  ) |>
  select(term, estimate_stars, std.error, p.value)

# Build gt table
model4_tidy |>
  gt() |>
  tab_header(
    title    = "Fixed Effects Estimation of NAFTA's Impact on RCA",
    subtitle = "Dependent variable: Log(RCA)"
  ) |>
  cols_label(
    term = "Variable",
    estimate_stars = "Coefficient",
    std.error = "Std. Error",
    p.value = "p-value"
  ) |>
  fmt_number(
    columns = p.value,
    decimals = 3
  ) |>
  tab_spanner(
    label = "Log(RCA)",
    columns = c(estimate_stars, std.error, p.value)
  ) |>
  tab_footnote(
    footnote = "*** p < 0.01, ** p < 0.05, * p < 0.1"  #shows statistical significance
  ) |>
  tab_footnote(
    footnote = "Clustered standard errors at the cmdCode level. Reference category: Pre-NAFTA."
  ) |>
  tab_source_note(
    source_note = "Source: UN Comtrade, Secretaría de Economía, Banco de México."
  ) |>
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    column_labels.font.weight  = "bold",
    table.width = pct(100)
  )

```


## Transition table

```{r}

#Generated by Claude

transition_table <- rca_panel |>
  filter(post_nafta == 0) |>
  group_by(cmdCode) |>
  summarise(had_ca = any(has_CA)) |>
  inner_join(
    rca_panel |>
      filter(post_nafta == 1) |>
      group_by(cmdCode) |>
      summarise(gained_ca = any(has_CA)),
    by = "cmdCode"
  ) |>
  mutate(transition = case_when(
    !had_ca &  gained_ca ~ "Gained RCA post-NAFTA",
    had_ca  & !gained_ca ~ "Lost RCA post-NAFTA",
    had_ca  &  gained_ca ~ "Maintained RCA",
    !had_ca & !gained_ca ~ "Never had RCA"
  )) |>
  count(transition) |>
  rename(
    `Transition Status` = transition,
    `Number of Products` = n
  )


transition_table |>
  gt() |>
  tab_header(
    title    = "Comparative Advantage Transitions Around NAFTA (1994)",
    subtitle = "Mexico's Agricultural Sector — HS Codes 01–53"
  ) |>
  tab_spanner(
    label   = "Pre vs. Post-NAFTA CA Status",
    columns = everything()
  ) |>
  cols_align(align = "left",  columns = `Transition Status`) |>
  cols_align(align = "center", columns = `Number of Products`) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = seq(1, nrow(transition_table), 2))
  ) |>
  tab_source_note(
    source_note = "Source: UN Comtrade. RCA calculated using Balassa (1965) index. CA defined as RCA ≥ 1."
  ) |>
  opt_table_font(font = "Times New Roman") |>  # standard for economics papers
  opt_stylize(style = 1, color = "gray")
```

# MACHINE LEARNING

We can use Synthetic Control at the sector level treating each HS2 sector as a unit, using other sectors as the "donor pool." The question this can answer would be for a given sector, what would its RCA trajectory have looked like if NAFTA had not happened, constructed as a weighted combination of other sectors?


```{r}

winsorize <- function(x, low = 0.01, high = 0.99) {
  q <- quantile(x, probs = c(low, high), na.rm = TRUE)
  pmax(pmin(x, q[2]), q[1])
}

# ── Prep: sector-level panel ──────────────────────────────────────────────────
sector_panel <- rca_panel |>
  filter(
    !is.na(RCA),
    is.finite(RCA),
    RCA > 0,
    nchar(cmdCode) == 4
  ) |>
  mutate(RCA_w = winsorize(RCA)) |>
  group_by(sector, period) |>
  summarise(
    log_mean_rca = mean(log(RCA_w), na.rm = TRUE),
    n_products   = n_distinct(cmdCode),
    .groups      = "drop"
  )


# which years does each sector actually cover? are there any years that 
sector_panel |>
  group_by(sector) |>
  summarise(
    min_year = min(period),
    max_year = max(period),
    n_years = n_distinct(period),
    missing = paste(setdiff(min_year:max_year, period), collapse = ", ")
  ) |>
  arrange(n_years)
```

```{r}
# ── Drop sectors with incomplete year coverage ────────────────────────────────
full_years <- sort(unique(sector_panel$period))

complete_sectors <- sector_panel |>
  group_by(sector) |>
  summarise(n_years = n_distinct(period)) |>
  filter(n_years == length(full_years)) |>
  pull(sector)

cat("Sectors retained:", length(complete_sectors), "\n")
cat("Sectors dropped: ", 
    paste(setdiff(unique(sector_panel$sector), complete_sectors), 
          collapse = ", "), "\n")

sector_panel_clean <- sector_panel |>
  filter(sector %in% complete_sectors)

# ── Synthetic control for sector 07 ──────────────────────────────────────────
synth_07 <- sector_panel_clean |>
  synthetic_control(
    outcome = log_mean_rca,
    unit = sector,
    time = period,
    i_unit  = "07",
    i_time = 1994,
    generate_placebos = TRUE
  ) |>
  generate_predictor(
    time_window = 1992,
    rca_1992    = log_mean_rca
  ) |>
  generate_predictor(
    time_window = 1993,
    rca_1993    = log_mean_rca
  ) |>
  generate_weights(
    optimization_window = 1990:1993,
    margin_ipop  = 0.02,
    sigf_ipop = 7,
    bound_ipop = 6
  ) |>
  generate_control()

# ── Plots ─────────────────────────────────────────────────────────────────────
plot_trends(synth_07) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  labs(
    title    = "Sector 07 (Vegetables): Actual vs Synthetic RCA",
    subtitle = "Synthetic donor pool = all other complete agricultural sectors",
    x = "Year", y = "log(mean RCA)"
  ) +
  theme_bw()

plot_differences(synth_07) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  labs(
    title    = "NAFTA Effect on Sector 07: Gap Between Actual and Synthetic",
    subtitle = "Positive = actual RCA exceeded synthetic counterfactual",
    x = "Year", y = "Δ log(mean RCA)"
  ) +
  theme_bw()

plot_placebos(synth_07) +
  labs(
    title    = "Placebo Tests: Sector 07 vs Donor Pool",
    subtitle = "Sector 07 line should stand out from placebo distribution",
    x = "Year", y = "Δ log(mean RCA)"
  ) +
  theme_bw()

plot_weights(synth_07) +
  labs(title = "Donor Sector Weights — Sector 07 Synthetic Control")

# ── Significance ──────────────────────────────────────────────────────────────
grab_significance(synth_07)
```


## DiD model

```{r, DiD DF}
## Clean DiD setup (NAFTA is a common time shock)

did_df <- rca_panel |>
  filter(!is.na(log_rca), is.finite(log_rca)) |>
  mutate(unit_id = as.numeric(factor(cmdCode))) |>
  arrange(unit_id, period)

## NAFTA post dummy (global shock)
did_df <- did_df |>
  mutate(post_nafta = as.integer(period >= 1994))

## Correct DiD / TWFE model (NO period FE, NO event_time)
es_model <- feols(
  log_rca ~ post_nafta + log(fdi_inflow_usd) + log_exch | unit_id,
  data = did_df,
  cluster = ~unit_id
)

summary(es_model)

```


After NAFTA

- about 0.58 log points lower on average, holding product fixed effects constant and controlling for macro variables.

- Post-NAFTA, revealed comparative advantage declines by ~44% within products relative to their pre-NAFTA baseline.

FDI Effect

- A 1% increase in FDI inflows is associated with approximately 0.18% decrease in RCA, holding other factors constant.


- FDI inflows may be associated with reduced revealed comparative advantage in domestic product production, possibly due to:

  -     foreign substitution of domestic production
  
  -     restructuring toward foreign-controlled production chains
  
-      reallocation away from traditional RCA sectors

Exchange rate

-   A 1% increase in exchange rate (depreciation depending on definition) is associated with ~0.23% increase in RCA, but not statistically reliable.

- weaker currency → exports become more competitive → RCA rises




## SCM Model

### Specification

The synthetic control estimator follows @abadie2010synthetic. Confidential notes from Dr. Gautam Sethi of Bard College was of usage as well. His textbook can be found here

-   (https://www.mit.edu/~jhainm/Paper/ccs.pdf)

-   (https://analytics.thephilomath.org/chapter/panel-data)

Let $Y_{it}$ denote the log RCA of unit $i$ in year $t$, 
where $i \in \{$Mexico, Spain, India, Indonesia, Thailand, 
Bolivia, Argentina, Guatemala, Morocco$\}$ and $t \in \{1990, \ldots, 2000\}$.

**Treated unit:** Mexico ($i = 1$), receiving treatment (NAFTA) at $t^* = 1994$.

**Donor pool:** $J$ countries unexposed to NAFTA, indexed $i = 2, \ldots, J+1$.

The synthetic control is a weighted combination of donor units:

$$
\hat{Y}_{1t}^{(0)} = \sum_{j=2}^{J+1} w_j^* Y_{jt}
$$

where $w_j^* \geq 0$ for all $j$ and $\sum_{j=2}^{J+1} w_j^* = 1$.

The optimal weights $\mathbf{W}^* = (w_2^*, \ldots, w_{J+1}^*)$ are chosen to minimize 
the pre-treatment discrepancy:

$$
\mathbf{W}^* = \arg\min_{\mathbf{W}} \sum_{m=1}^{k} v_m \left( X_{1m} - \sum_{j=2}^{J+1} w_j X_{jm} \right)^2
$$

subject to $w_j \geq 0$ and $\sum_{j=2}^{J+1} w_j = 1$, where $X_{im}$ denotes 
the $m$-th pre-treatment predictor for unit $i$, specifically, 
$\log(\text{RCA})$ in years 1990, 1991, 1992, and 1993, 
as well as the mean pre-treatment $\log(\text{RCA})$ over $1990$--$1993$ 
and $v_m$ are predictor weights chosen to minimize out-of-sample 
prediction error over the pre-treatment period.

The estimated treatment effect of NAFTA on Mexico's vegetable RCA in year $t \geq 1994$ is:

$$
\hat{\tau}_{1t} = Y_{1t} - \hat{Y}_{1t}^{(0)} = Y_{1t} - \sum_{j=2}^{J+1} w_j^* Y_{jt}
$$

Statistical inference is conducted via the permutation-based placebo test of 
@abadie2010synthetic. Each donor country $j$ is iteratively assigned as the 
placebo treated unit, and a synthetic control is constructed from the remaining 
donors. The post- to pre-treatment mean squared prediction error (MSPE) ratio is 
computed for each unit:

$$
\text{MSPE Ratio}_i = \frac{\frac{1}{T_1}\sum_{t \geq t^*} (Y_{it} - \hat{Y}_{it}^{(0)})^2}
{\frac{1}{T_0}\sum_{t < t^*} (Y_{it} - \hat{Y}_{it}^{(0)})^2}
$$

where $T_0$ and $T_1$ denote the number of pre- and post-treatment periods 
respectively. The Fisher's exact $p$-value is the proportion of units 
(including Mexico) with an MSPE ratio at least as large as Mexico's.



### Windsorization

```{r}
country_a_xports_sector_07 <- read_csv(
  "regression/scm2/country_a_xports_sector_07.csv",
  locale = locale(encoding = "Latin1")
) |>
  mutate(reporterDesc = stringi::stri_trans_general(reporterDesc, "Latin-ASCII"))

country_a_xports_of_all_prods <- read_csv(
  "regression/scm2/country_a_xports_of_all_prods.csv",
  locale = locale(encoding = "Latin1")
) |>
  mutate(reporterDesc = stringi::stri_trans_general(reporterDesc, "Latin-ASCII"))

```

```{r}

# ── Winsorize ────────────────────────────────────────────────────────────────
winsorize <- function(x, low = 0.01, high = 0.99) {
  q <- quantile(x, probs = c(low, high), na.rm = TRUE)
  pmax(pmin(x, q[2]), q[1])
}

years <- 1990:2000

# ── Step 1: World aggregates ─────────────────────────────────────────────────
world_prod_07 <- map_dfr(time_periods, function(t) {
  world_file <- file.path(base_path,
                          paste0("world_xports_prod_i_time_", t),
                          paste0("world_xports_sector_07_time_", t, ".csv"))
  if (!file.exists(world_file)) return(NULL)

  read_csv(world_file, show_col_types = FALSE) |> necessary_variables()
}) |>
  filter(period %in% years) |>
  group_by(period) |>
  summarise(world_product_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

world_total <- world_xports_all |>
  group_by(period) |>
  summarise(total_world_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop") |>
  filter(period %in% years)

# ── Step 2: Country-level exports ────────────────────────────────────────────
donor_prod <- country_a_xports_sector_07 |>
  group_by(reporterDesc, period) |>
  summarise(country_product_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

donor_total <- country_a_xports_of_all_prods |>
  group_by(reporterDesc, period) |>
  summarise(total_country_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

# ── Step 3: RCA calculation ──────────────────────────────────────────────────
donor_rca <- donor_prod |>
  left_join(donor_total,   by = c("reporterDesc", "period")) |>
  left_join(world_prod_07, by = "period") |>
  left_join(world_total,   by = "period") |>
  mutate(
    RCA = (country_product_value / total_country_value) /
          (world_product_value   / total_world_value),
    RCA_w = winsorize(RCA),
    RCA_w = if_else(RCA_w <= 0, NA_real_, RCA_w),   # 🔑 critical fix
    log_mean_rca = log(RCA_w),
    country = reporterDesc
  ) |>
  select(country, period, log_mean_rca)

# ── Step 4: Mexico RCA ───────────────────────────────────────────────────────
mexico_07 <- rca_big_df |>
  filter(sector == "07",
         nchar(cmdCode) == 4,
         period %in% years) |>
  mutate(
    RCA_w = winsorize(RCA),
    RCA_w = if_else(RCA_w <= 0, NA_real_, RCA_w)
  ) |>
  group_by(period) |>
  summarise(log_mean_rca = mean(log(RCA_w), na.rm = TRUE),
            .groups = "drop") |>
  mutate(country = "Mexico")

# ── Step 5: FORCE BALANCED PANEL ─────────────────────────────────────────────
donor_rca_balanced <- donor_rca |>
  complete(country, period = years)

mexico_07 <- mexico_07 |>
  complete(period = years)

# ── Step 6: Drop countries with ANY missing data ─────────────────────────────
valid_countries <- donor_rca_balanced |>
  group_by(country) |>
  summarise(
    n_years = sum(!is.na(log_mean_rca)),
    has_na= any(is.na(log_mean_rca) | !is.finite(log_mean_rca))
  ) |>
  filter(n_years == length(years), !has_na) |>
  pull(country)

donor_rca_clean <- donor_rca_balanced |>
  filter(country %in% valid_countries)

# Check Mexico validity
if (any(is.na(mexico_07$log_mean_rca))) {
  stop("Mexico has missing data — cannot run SCM")
}

# ── Step 7: Final panel ──────────────────────────────────────────────────────
donor_panel <- bind_rows(mexico_07, donor_rca_clean)

# Diagnostics

cat("Final donor countries:\n")

donor_panel |> distinct(country) |> print() #shows us the donor countries with data from 1990:2000

donor_panel <- donor_panel |>
  group_by(country, period) |>
  summarise(log_mean_rca = mean(log_mean_rca, na.rm = TRUE),
            .groups = "drop") |>
  filter(country != "Cyprus") 


# ── Step 8: Pre-treatment plot ───────────────────────────────────────────────
ggplot(donor_panel,
       aes(x = period, y = log_mean_rca, color = country,
           linewidth = if_else(country == "Mexico", 1.2, 0.6))) +
  geom_line() +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  scale_linewidth_identity() +
  labs(title = "Pre-treatment trends (1990–1993)",
       x = "Year", y = "log(RCA)") +
  theme_bw()

# ── Step 9: Synthetic Control ────────────────────────────────────────────────
synth_mexico <- donor_panel |>
  synthetic_control(
    outcome = log_mean_rca,
    unit= country,
    time = period,
    i_unit = "Mexico",
    i_time = 1994,
    generate_placebos = TRUE
  ) |>
  generate_predictor(time_window = 1990:1993,
                     mean_pre = mean(log_mean_rca, na.rm = TRUE)) |>
  generate_predictor(time_window = 1990, rca_1990 = log_mean_rca) |>
  generate_predictor(time_window = 1991, rca_1991 = log_mean_rca) |>
  generate_predictor(time_window = 1992, rca_1992 = log_mean_rca) |>
  generate_predictor(time_window = 1993, rca_1993 = log_mean_rca) |>
  generate_weights(
    optimization_window = 1990:1993,
    margin_ipop = 0.02,
    sigf_ipop = 7,
    bound_ipop = 6
  ) |>
  generate_control()

# ── Step 10: Plots ───────────────────────────────────────────────────────────
plot_trends(synth_mexico) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  labs(title = "Actual vs Synthetic Mexico",
       y = "log(RCA)") +
  theme_bw()

plot_differences(synth_mexico) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  labs(title = "Treatment Effect (Gap)",
       y = "Δ log(RCA)") +
  theme_bw()

plot_placebos(synth_mexico) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue")

plot_weights(synth_mexico)

# ── Step 11: Inference ───────────────────────────────────────────────────────
grab_significance(synth_mexico)

```

### Attempt 4 Non - Wind

```{r}
library(tidyverse)
library(tidysynth)

# ─────────────────────────────────────────────────────────────
# PARAMETERS
# ─────────────────────────────────────────────────────────────
years <- 1990:2000
eps <- 1e-6
# ─────────────────────────────────────────────────────────────
# WORLD DATA
# ─────────────────────────────────────────────────────────────
world_prod_07 <- map_dfr(time_periods, function(t) {
  world_file <- file.path(base_path,
                          paste0("world_xports_prod_i_time_", t),
                          paste0("world_xports_sector_07_time_", t, ".csv"))
  if (!file.exists(world_file)) return(NULL)

  read_csv(world_file, show_col_types = FALSE) |>
    necessary_variables()
}) |>
  filter(period %in% years) |>
  group_by(period) |>
  summarise(world_product_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

world_total <- world_xports_all |>
  group_by(period) |>
  summarise(total_world_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop") |>
  filter(period %in% years)

# ─────────────────────────────────────────────────────────────
# DONOR DATA
# ─────────────────────────────────────────────────────────────
donor_prod <- country_a_xports_sector_07 |>
  group_by(reporterDesc, period) |>
  summarise(country_product_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

donor_total <- country_a_xports_of_all_prods |>
  group_by(reporterDesc, period) |>
  summarise(total_country_value = sum(fobvalue, na.rm = TRUE),
            .groups = "drop")

# ─────────────────────────────────────────────────────────────
# NEW STABLE RCA (NO WINSORIZATION)
# ─────────────────────────────────────────────────────────────
donor_rca <- donor_prod |>
  left_join(donor_total,   by = c("reporterDesc", "period")) |>
  left_join(world_prod_07, by = "period") |>
  left_join(world_total,   by = "period") |>
  mutate(
    log_RCA =
      log(country_product_value + eps) -
      log(total_country_value + eps) -
      (log(world_product_value + eps) -
       log(total_world_value + eps)),
    country = reporterDesc
  ) |>
  select(country, period, log_RCA)

# ─────────────────────────────────────────────────────────────
# MEXICO (same transformation for consistency)
# ─────────────────────────────────────────────────────────────
mexico_07 <- rca_big_df |>
  filter(sector == "07",
         nchar(cmdCode) == 4,
         period %in% years) |>
  group_by(period) |>
  summarise(
    log_RCA = mean(
      log((RCA + eps)),
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(country = "Mexico")

# ─────────────────────────────────────────────────────────────
# BALANCE PANEL (CRITICAL FOR SCM)
# ─────────────────────────────────────────────────────────────
donor_panel <- bind_rows(mexico_07, donor_rca) |>
  group_by(country, period) |>
  summarise(log_RCA = mean(log_RCA, na.rm = TRUE),
            .groups = "drop") |>
  complete(country, period = years)

# ─────────────────────────────────────────────────────────────
# DROP INCOMPLETE COUNTRIES (NO MISSING YEARS)
# ─────────────────────────────────────────────────────────────
valid_countries <- donor_panel |>
  filter(country != "Cyprus") |>
  group_by(country) |>
  summarise(
    ok = all(!is.na(log_RCA) & is.finite(log_RCA)),
    n  = sum(!is.na(log_RCA))
  ) |>
  filter(ok, n == length(years)) |>
  pull(country)

donor_panel <- donor_panel |>
  filter(country %in% valid_countries)

# ─────────────────────────────────────────────────────────────
# SCM MODEL
# ─────────────────────────────────────────────────────────────
synth_mexico <- donor_panel |>
  synthetic_control(
    outcome = log_RCA,
    unit  = country,
    time = period,
    i_unit = "Mexico",
    i_time = 1994,
    generate_placebos = TRUE
  ) |>
  generate_predictor(time_window = 1990:1993,
                     mean_pre = mean(log_RCA, na.rm = TRUE)) |>
  generate_predictor(time_window = 1990, rca_1990 = log_RCA) |>
  generate_predictor(time_window = 1991, rca_1991 = log_RCA) |>
  generate_predictor(time_window = 1992, rca_1992 = log_RCA) |>
  generate_predictor(time_window = 1993, rca_1993 = log_RCA) |>
  generate_weights(
    optimization_window = 1990:1993,
    margin_ipop = 0.02,
    sigf_ipop = 7,
    bound_ipop = 6
  ) |>
  generate_control()

# ─────────────────────────────────────────────────────────────
# PLOTS
# ─────────────────────────────────────────────────────────────
plot_trends(synth_mexico) +
  geom_vline(xintercept = 1994, linetype = "dashed", color = "blue")

plot_differences(synth_mexico) +
 geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  annotate("text", x = 1994.2, y = Inf, vjust = 1.5,
           hjust = 0, size = 3, color = "blue", label = "NAFTA") +
  labs(title    = "NAFTA Effect: Actual minus Synthetic log(RCA)",
       subtitle = "Sector 07 Vegetables — negative gap = below counterfactual",
       x = "Year", y = "Δ log(RCA)") +
  theme_bw()

plot_placebos(synth_mexico) +
 geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
  labs(title    = "Placebo Tests: Mexico vs Donor Countries",
       subtitle = "Mexico's gap should be larger than donor country gaps",
       x = "Year", y = "Δ log(RCA)") +
  theme_bw()

plot_weights(synth_mexico)

# ─────────────────────────────────────────────────────────────
# INFERENCE
# ─────────────────────────────────────────────────────────────
k <- grab_significance(synth_mexico)
```

### Interpretationof SCM

Placebo Plot (Image 1)

Mexico's pink line is persistently below zero post-1994 and stays there through 2000, while most donor country gray lines oscillate around zero. This is exactly what you want to see — Mexico's gap is distinctive and sustained. However, some donor lines also dip negative post-1994, which means Mexico's gap is not uniquely extreme. This matters for the significance test below.

Trends Plot (Image 2)
Similarly as previously, there exists persistent ~0.7 log point gap post-1994, while synthetic stays flat around 1.9 while Mexico stays around 1.1. The finding is robust to dropping Cyprus, as Cyprus is a small economy with pre-treatment behavior exhibiting a trend similar to Mexico skewing the weights.

Significance Table (Image 3)

Mexico's rank is 18 out of 23. The MSPE ratio (post/pre mean squared prediction error) for Mexico is 1.72. This means Mexico's post-treatment fit is only 1.72 times worse than its pre-treatment fit which is a low ratio. Seventeen donor countries have a higher MSPE ratio than Mexico, meaning their post-treatment gaps are larger relative to their pre-treatment fit than Mexico's is.

The Fisher's exact p-value is 0.78. This is not statistically significant by any conventional threshold. You cannot reject the null hypothesis that NAFTA had no effect on Mexico's vegetable RCA.


Hence, the synthetic control estimates suggest Mexico's log(RCA) in the selected HS codes for sector 07 was approximately 0.7 log points below its synthetic counterfactual throughout 1994–2000. This is consistent with the FE estimates in Table X. However, the pre-treatment fit is poor as the pre-MSPE = 0.433, reflecting macroeconomic volatility in Mexico during 1990–1993 that the donor pool countries are unable to replicate replicate. As a result, Mexico's post/pre MSPE ratio of 1.72 ranks 18th out of 23 units, ultimately yielding a Fisher's p-value of 0.78. The synthetic control therefore provides suggestive but rather statistically inconclusive evidence of a NAFTA effect on Mexico's vegetable comparative advantage. These results should be interpreted alongside the panel regression estimates as corroborating descriptive evidence rather than causal identification.

For next steps, this analysis can be done for other sectors, like the Vehicles and Machinery sector, and 



# Miscellaneous and Extra Code

```{r}
##GENERALIZED RCA ANALYSIS

## Generalized RCA function for any product code
#calculate_rca_for_product <- function(product_code) {
  
  #  build file paths for  product
 # mx_prod_path <- paste0("mx_xports_prod_i/mx_xports_", product_code, ".csv")
 # world_prod_path <- paste0("world_xports_prod_i/world_xports_", product_code, ".csv")
  
  # load and read the  data
  #mx_xports_prod <- read_csv(mx_prod_path, show_col_types = FALSE)
 # world_xports_prod <- read_csv(world_prod_path, show_col_types = FALSE)
  
  # laod world totals for time frame 
  #mx_xports_all <- read_csv("mx_total_xports_world/mx_xports_all.csv", show_col_types = FALSE)
  #world_xports_all <- read_csv("world_total_xports/world_xports_all.csv", show_col_types = FALSE)
  
  # clean and arrange using  necessary_varioables function
#  mx_xports_prod <- necessary_variables(mx_xports_prod)
#  world_xports_prod <- necessary_variables(world_xports_prod)
#  mx_xports_all <- necessary_variables(mx_xports_all)
#  world_xports_all <- necessary_variables(world_xports_all)
  
  # calculate RCA
#  rca_df <- mx_xports_prod %>%
 #   inner_join(mx_xports_all, by = "period", suffix = c("_prod", "_total")) %>%
#    inner_join(world_xports_prod, by = "period") %>%
 #   inner_join(world_xports_all, by = "period", suffix = c("_world_prod", "_world_total")) %>%
 #   mutate(
   #   RCA = (fobvalue_prod / fobvalue_total) /
  #          (fobvalue_world_prod / fobvalue_world_total)
 #   ) %>%
  #  select(period, RCA) %>%
 #   mutate(product_code = product_code)
  
 # return(rca_df)
#}
```

### run RCA all sectors

```{r}
rca_all <- purrr::map_dfr(sector_codes, process_sector)

## keep only first two digits

rca_all <- rca_all |>
  mutate(
    cmdCode = as.character(cmdCode),
    sector = substr(cmdCode, 1, 2)
  )
```


```{r}
sectors <- unique(rca_all$sector) ## creatrs a list for all elements in sector_code MAKE SURE TO ADD EACH SECTOR CODE!


for (s in sectors) { ## for all elements s in list sectors,
  
  allplots <- ggplot( 
    rca_all |> filter(sector == s), ## filter for HS code loops
    aes(x = period, y = RCA, color = cmdCode, group = cmdCode)
  ) +
    geom_line(size = 1) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    labs(
      title = paste("RCA for Sector", s),
      x = "Year",
      y = "RCA",
      color = "HS Product"
    ) +
    theme_minimal()
  
  print(allplots)
}
```


### SCM Failed Attempts

### Attempt 1

```{r}
# 
# library(tidysynth)
# 
# # ── Winsorize function ────────────────────────────────────────────────────────
# winsorize <- function(x, low = 0.01, high = 0.99) {
#   q <- quantile(x, probs = c(low, high), na.rm = TRUE)
#   pmax(pmin(x, q[2]), q[1])
# }
# 
# # ── Step 1: Load and clean country names ──────────────────────────────────────
# country_a_xports_prod_i <- read_csv("regression/scm/country_a_xports_prod_i.csv") |>
#   mutate(reporterDesc = case_when(
#     str_detect(reporterDesc, "rkiye") ~ "Turkey",
#     str_detect(reporterDesc, "Bolivia") ~ "Bolivia",
#     TRUE ~ reporterDesc
#   ))
# 
# country_a_xports_allprods <- read_csv("regression/scm/country_a_xports_allprods.csv") |>
#   mutate(reporterDesc = case_when(
#     str_detect(reporterDesc, "rkiye") ~ "Turkey",
#     str_detect(reporterDesc, "Bolivia") ~ "Bolivia",
#     TRUE ~ reporterDesc
#   ))
# 
# # ── Step 2: World sector 07 exports from existing pipeline files ─────────────
# world_prod_07 <- map_dfr(time_periods, function(t) {
#   world_file <- file.path(base_path,
#                           paste0("world_xports_prod_i_time_", t),
#                           paste0("world_xports_sector_07_time_", t, ".csv"))
#   if (!file.exists(world_file)) {
#     message("Missing: ", world_file)
#     return(NULL)
#   }
#   read_csv(world_file, show_col_types = FALSE) |> necessary_variables()
# }) |>
#   filter(period >= 1992, period <= 2000) |>
#   group_by(period) |>
#   summarise(world_product_value = sum(fobvalue, na.rm = TRUE),
#             .groups = "drop")
# 
# # World total exports
# world_total <- world_xports_all |>
#   group_by(period) |>
#   summarise(total_world_value = sum(fobvalue, na.rm = TRUE),
#             .groups = "drop") |>
#   filter(period >= 1992, period <= 2000)
# 
# # ── Step 3: Compute donor country RCAs ───────────────────────────────────────
# 
# donor_prod <- country_a_xports_prod_i |>
#   group_by(reporterDesc, period) |>
#   summarise(country_product_value = sum(fobvalue, na.rm = TRUE),
#             .groups = "drop")
# 
# donor_total <- country_a_xports_allprods |>
#   group_by(reporterDesc, period) |>
#   summarise(total_country_value = sum(fobvalue, na.rm = TRUE),
#             .groups = "drop")
# 
# donor_rca <- donor_prod |>
#   left_join(donor_total,   by = c("reporterDesc", "period")) |>
#   left_join(world_prod_07, by = "period") |>
#   left_join(world_total,   by = "period") |>
#   mutate(
#     RCA = (country_product_value / total_country_value) /
#                    (world_product_value   / total_world_value),
#     RCA_w = winsorize(RCA),
#     log_mean_rca = log(RCA_w),
#     country = reporterDesc
#   ) |>
#   select(country, period, log_mean_rca)
# 
# # ── Step 4: Mexico sector 07 RCA ─────────────────────────────────────────────
# 
# mexico_07 <- rca_big_df |>
#   filter(sector == "07",
#          nchar(cmdCode) == 4,
#          period >= 1992,
#          period <= 2000) |>
#   mutate(RCA_w = winsorize(RCA)) |>
#   group_by(period) |>
#   summarise(log_mean_rca = mean(log(RCA_w), na.rm = TRUE),
#             .groups = "drop") |>
#   mutate(country = "Mexico")
# 
# # ── Step 5: Stack and clean donor panel ───────────────────────────────────────
# 
# drop_countries <- c("Kenya", "Costa Rica", "Honduras", "Turkey", "Ecuador")
# 
# donor_panel <- bind_rows(mexico_07, donor_rca) |>
#   filter(
#     !country %in% drop_countries,
#     period >= 1992,
#     period <= 2000,
#     !is.na(log_mean_rca),
#     is.finite(log_mean_rca)
#   )
# 
# # ── Step 6: Completeness check ────────────────────────────────────────────────
# 
# cat("Countries in donor pool:\n")
# donor_panel |> distinct(country) |> print()
# 
# cat("\nCompleteness matrix:\n")
# donor_panel |>
#   pivot_wider(names_from = country, values_from = log_mean_rca) |>
#   arrange(period) |>
#   print(n = 9)
# 
# # ── Step 7: Pre-treatment trajectory plot ──────
# ggplot(donor_panel,
#        aes(x = period, y = log_mean_rca, color = country,
#            linewidth = if_else(country == "Mexico", 1.2, 0.6))) +
#   geom_line() +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   scale_linewidth_identity() +
#   labs(title    = "Sector 07 log(RCA): Mexico and Donor Countries",
#        subtitle = "Pre-1994 parallel trends check",
#        x = "Year", y = "log(mean RCA)", color = "Country") +
#   theme_bw()
# 
# 
# # ── Step 8: Synthetic control ─────────────────────────────────────────────────
# synth_mexico <- donor_panel |>
#   synthetic_control(
#     outcome           = log_mean_rca,
#     unit              = country,
#     time              = period,
#     i_unit            = "Mexico",
#     i_time            = 1994,
#     generate_placebos = TRUE
#   ) |>
#   generate_predictor(time_window = 1992:1993,
#                      mean_pre = mean(log_mean_rca, na.rm = TRUE)) |>
#   generate_predictor(time_window = 1992, rca_1992 = log_mean_rca) |>
#   generate_predictor(time_window = 1993, rca_1993 = log_mean_rca) |>
#   generate_weights(
#     optimization_window = 1992:1993,
#     margin_ipop         = 0.02,
#     sigf_ipop           = 7,
#     bound_ipop          = 6
#   ) |>
#   generate_control()
# 
# # ── Step 9: Plots ─────────────────────────────────────────────────────────────
# 
# # Actual vs synthetic
# 
# plot_trends(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   labs(title    = "Mexico Sector 07 (Vegetables): Actual vs Synthetic RCA",
#        subtitle = "Donor pool: Indonesia, Spain, Bolivia, Ecuador, Morocco, Guatemala, Argentina",
#        x = "Year", y = "log(mean RCA)") +
#   theme_bw()
# 
# # Gap plot
# plot_differences(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   annotate("text", x = 1994.1, y = Inf, vjust = 1.5,
#            hjust = 0, size = 3, color = "blue", label = "NAFTA") +
#   labs(title    = "NAFTA Effect: Gap Between Actual and Synthetic log(RCA)",
#        subtitle = "Positive = Mexico outperformed its counterfactual",
#        x = "Year", y = "Δ log(mean RCA)") +
#   theme_bw()
# 
# # Placebo tests
# plot_placebos(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   labs(title    = "Placebo Tests: Mexico vs Donor Countries",
#        subtitle = "Mexico's post-1994 gap should exceed placebo distribution",
#        x = "Year", y = "Δ log(mean RCA)") +
#   theme_bw()
# 
# # Donor weights
# plot_weights(synth_mexico) +
#   labs(title    = "Synthetic Control Donor Weights",
#        subtitle = "Which countries construct Mexico's counterfactual")
# 
# # ── Step 10: Significance ─────────────────────────────────────────────────────
# grab_signifiance(synth_mexico)

```

### Attempt 2

```{r}
#library(tidysynth)

# ── Winsorize function (if not already defined) ───────────────────────────────
#winsorize <- function(x, low = 0.01, high = 0.99) {
 # q <- quantile(x, probs = c(low, high), na.rm = TRUE)
  #pmax(pmin(x, q[2]), q[1])
#}

# ── Step 1: World sector 07 exports from existing pipeline files ──────────────
#world_prod_07 <- map_dfr(time_periods, function(t) {
 # world_file <- file.path(base_path,
  #                        paste0("world_xports_prod_i_time_", t),
   #                       paste0("world_xports_sector_07_time_", t, ".csv"))
  #if (!file.exists(world_file)) {
   # message("Missing: ", world_file)
 #   return(NULL)
 # }
 # read_csv(world_file, show_col_types = FALSE) |> necessary_variables()
#}) |>
 # filter(period >= 1990, period <= 2000) |>
#  group_by(period) |>
 # summarise(world_product_value = sum(fobvalue, na.rm = TRUE),
  #          .groups = "drop")

# World total exports (aggregate — cmdCode = -2)
#world_total <- world_xports_all |>
#  group_by(period) |>
 # summarise(total_world_value = sum(fobvalue, na.rm = TRUE),
  #          .groups = "drop") |>
 # filter(period >= 1990, period <= 2000)

# ── Step 2: Donor country RCAs ────────────────────────────────────────────────
#donor_prod <- scm_countries_xports_07 |>
#  group_by(reporterDesc, period) |>
 # summarise(country_product_value = sum(fobvalue, na.rm = TRUE),
      #      .groups = "drop")

#donor_total <- scm_countries_xports_all |>
#  group_by(reporterDesc, period) |>
 # summarise(total_country_value = sum(fobvalue, na.rm = TRUE),
   #         .groups = "drop")

#donor_rca <- donor_prod |>
 # left_join(donor_total,   by = c("reporterDesc", "period")) |>
 # left_join(world_prod_07, by = "period") |>
 # left_join(world_total,   by = "period") |>
  #mutate(
  #  RCA          = (country_product_value / total_country_value) /
  #                 (world_product_value   / total_world_value),
  #  RCA_w        = winsorize(RCA),
   # log_mean_rca = log(RCA_w),
  #  country      = reporterDesc
#  ) |>
  #select(country, period, log_mean_rca)

# ── Step 3: Mexico sector 07 RCA ─────────────────────────────────────────────
#mexico_07 <- rca_big_df |>
 # filter(sector == "07",
  #       nchar(cmdCode) == 4,
   #      period >= 1990,
   #      period <= 2000) |>
 # mutate(RCA_w = winsorize(RCA)) |>
#  group_by(period) |>
 # summarise(log_mean_rca = mean(log(RCA_w), na.rm = TRUE),
    #        .groups = "drop") |>
 # mutate(country = "Mexico")

# ── Step 4: Stack into donor panel ───────────────────────────────────────────
#donor_panel <- bind_rows(mexico_07, donor_rca)

# Sanity check
#onor_panel |> count(country)
#summary(donor_panel$log_mean_rca)

# ── Step 5: Pre-treatment trajectory plot ────────────────────────────────────
# Visually check whether donor countries track Mexico pre-1994
#ggplot(donor_panel, 
 #      aes(x = period, y = log_mean_rca, color = country,
  #         linewidth = if_else(country == "Mexico", 1.2, 0.6))) +
 # geom_line() +
 # geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
# scale_linewidth_identity() +
#  labs(title    = "Sector 07 log(RCA): Mexico and Donor Countries",
#       subtitle = "Pre-1994 parallel trends check",
 #      x = "Year", y = "log(mean RCA)", color = "Country") +
 # theme_bw()

# ── Step 6: Synthetic control ─────────────────────────────────────────────────
#synth_mexico <- donor_panel |>
 # synthetic_control(
 #   outcome           = log_mean_rca,
 #   unit              = country,
 #   time              = period,
 #   i_unit            = "Mexico",
#    i_time            = 1994,
##    generate_placebos = TRUE
#   ) |>
#   generate_predictor(time_window = 1990, rca_1990 = log_mean_rca) |>
#   generate_predictor(time_window = 1991, rca_1991 = log_mean_rca) |>
#   generate_predictor(time_window = 1992, rca_1992 = log_mean_rca) |>
#   generate_predictor(time_window = 1993, rca_1993 = log_mean_rca) |>
#   generate_weights(
#     optimization_window = 1990:1993,
#     margin_ipop         = 0.02,
#     sigf_ipop           = 7,
#     bound_ipop          = 6
#   ) |>
#   generate_control()
# 
# # ── Step 7: Plots ─────────────────────────────────────────────────────────────
# 
# # Actual vs synthetic trajectory
# plot_trends(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   labs(title    = "Mexico Sector 07 (Vegetables): Actual vs Synthetic RCA",
#        subtitle = "Donor pool: Brazil, Chile, Colombia, Peru",
#        x = "Year", y = "log(mean RCA)") +
#   theme_bw()
# 
# # Gap plot: actual minus synthetic
# plot_differences(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   annotate("text", x = 1994.2, y = Inf, vjust = 1.5,
#            hjust = 0, size = 3, color = "blue", label = "NAFTA") +
#   labs(title    = "NAFTA Effect: Gap Between Actual and Synthetic log(RCA)",
#        subtitle = "Positive = Mexico outperformed its counterfactual",
#        x = "Year", y = "Δ log(mean RCA)") +
#   theme_bw()
# 
# # Placebo tests
# plot_placebos(synth_mexico) +
#   geom_vline(xintercept = 1994, linetype = "dashed", color = "blue") +
#   labs(title    = "Placebo Tests: Mexico vs Donor Countries",
#        subtitle = "Mexico's post-1994 gap should exceed placebo distribution",
#        x = "Year", y = "Δ log(mean RCA)") +
#   theme_bw()
# 
# # Donor weights
# plot_weights(synth_mexico) +
#   labs(title    = "Synthetic Control Weights",
#        subtitle = "Which countries construct Mexico's counterfactual")
# 
# # ── Step 8: Significance ──────────────────────────────────────────────────────
# grab_signifiance(synth_mexico)
```
